\documentclass[../main.tex]{subfiles}
\begin{document}
    \chapter*{Preface}\label{ch:preface}
    \addcontentsline{toc}{chapter}{Preface}
    
    This thesis could not have been completed without the help and support of various people. 
    In this section I want to thank people who helped me during the process by mentioning their name and describe how they helped me. 
    I start with a short story on how this research came to be.
    
    \paragraph{}
    The initial idea for this thesis was to dive into the topic of software analysis in order to find vulnerability in PHP programs.
    When diving further in this topic my interest in statical analysis grew, and I wanted to reconstruct data flows in a program to be able to perform taint analysis.
    In order to perform a dataflow analysis for object oriented programs written in PHP, the types of variables need to be known.
    This topic of resolving the run-time type at compile time was big enough to have a thesis on its own.
    
    % Jurgen Vinju
    \paragraph{}
    First of all I would like to thank Jurgen Vinju whoem helped me throughout the whole thesis process.
    I want to thank Jurgen for his endless enthusiasm, number of ideas and suggestions and personal help on coaching and mentoring me.
    Jurgen helped defining a research topic and gave me many directions where to go or look for whenever I got stuck on a tough subject.
    
    % Mark Hills: previous work on php analysis
    \paragraph{}
    Next I would like to thank Mark Hills.
    The implementation of this research is build on top of the PHP Analysis in Rascal (PHP AiR) framework, created by Mark.
    We used the framework to parse PHP files to Rascal ASTs and implement an $M^3$ model for PHP programs.
    
    % Bas Basten: samen gewerkt aan M3
    \paragraph{}
    I want to thank Bas Basten for the collaboration on improving and extending the $M^3$ model.
    After I made an initial version of the $M^3$ model for PHP, Bas help to improved the model by adding more facts to the model.
    With his expertise in Rascal and and my knowledge of PHP we've formed a very good team.
    
    % Winfred Peereboom: Mogelijkheid geboden om af te studeren
    \paragraph{}
    I would like to thank Winfred Peereboom for giving me the opportunity to write my thesis at Werkspot.
    During the period Winfred helped me on various aspects of coaching and mentoring me.
	
	% Girlfriend for endless support and patience.
    % All others that I forgot to mention      
    \paragraph{}
    Finally I would like to thank my girlfriend for her endless support and patience.
    And of course I want to thank everyone I forgot to mention here personally, but did help me direct or indirect, consciousness or subconsciousness.

\end{document}